A family of accessibility measures derived from spatial interaction principles

### 2025-05-23 paper/manuscript.qmd
Cleaned up a few small equation errors - and posted preprint link to repo page.

---
title: "A family of accessibility measures derived from spatial interaction principles"
subtitle: "Submitted preprint for peer-review (May 2025)"
author:
  - name: Anastasia Soukhov*
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Rafael H. M. Pereira
    email: rafael.pereira@ipea.gov.br
    affiliations:
        - name: Institute for Applied Economic Research - Ipea
          city: Brasília, Brazil
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliations:
        - name: University of Toronto Scarborough
          department: Department of Human Geography
          city: Toronto, Canada
    attributes:
        corresponding: true
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format:
  pdf:
    keep-tex: true
    documentclass: article
    header-includes: |
      \usepackage{draftwatermark} 
      \SetWatermarkText{DRAFT}
      \SetWatermarkScale{1.5}
      \SetWatermarkColor[gray]{0.95}
      \usepackage{float}
      \floatplacement{table}{H}
      \floatplacement{figure}{H}
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
#includes:
#      in_header: 'gt_packages.sty'
      #\usepackage{float}

@@

  *Email of corresponding author for preprint: soukhoa@mcmaster.ca

@@

Understanding $S_i$ as the inverse of Wilson's $A_i$ does not uncover any new meaning for $S_i$ itself. Indeed, mathematically it is true, $A_i$'s role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ i.e., keeping units balanced and proportionality based on constraints. Understanding $S_i$ itself as a balancing factor is not wholly helpful as Hansen and (Stewart before him) defined accessibility as a partial sum of the demographic force $F$ or the system-wide population potential of opportunities for spatial interaction (i.e., @q-stewart-population-potential-sum representing $V_i = \sum_j \frac{M_j}{d_{ij}}$ after losing $G$).

@@

Firstly, @tbl-small-system-land-use summarises the population (in units of 10,000s of people) and the opportunities (the number of physicians) per zone. [^1]. Considering @tbl-small-system-land-use's values, the Provider-to-Population-Ratio (PPR) in this system is `r sum(LU$D_j) / sum(LU$O_i)`. For reference, the number of physicians per 10,000 in Canada in 2022 was 24.97 [@whoMedicalDoctors102025]. Secondly, to pair the zonal population and opportunity information, the assumed cost of movement (in minutes of travel time) between origins and destinations is as shown in @tbl-small-system-cost.

@@

Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the proportional allocation factor for the total constrained opportunity case $\kappa_i^T$ is dimensionless:

$$
\kappa_i^T = \frac{\sum_j D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
$$

\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \kappa_i^T\sum_j W^{(2)}_j$). The role of $\kappa_i^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so they represent a proportion of the total number of opportunities in the region. $\kappa_i^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$. This is why we refer to $\kappa_i^T$ as a proportional allocation factor. On the other hand, the proportionality constant $K^T$ balances the units of $V^0_i$, the Hansen-type accessibility values, and is an alternative expression of the total constrained accessibility measure.

@@

Using the calculated proportionality constants for all zones and multiplying them by the unconstrained accessibility value $V^0_i$, the total opportunity constrained accessibility values for all zones and different travel behaviour scenarios is presented in @tbl-simple-example-total-opportunity-accessibility. $\kappa_i^T$ for each zone is not shown, but can be understood to be the unitless proportion of opportunities (of the total opportunities) allocated to each zone based on travel impedance.

```{r}
#| label: tbl-simple-example-total-opportunity-accessibility
#| tbl-cap: "Simple system: total opportunity constrained accessibility."

@@

In contrast to unconstrained accessibility $V^0_i$, imposing a constraint -in this case a total opportunity constraint for this variant- allows for the comparison of differences and ratios between regions and across different travel behaviour scenarios as well. Each value is effectively in units of physicians, with the impedance units already accounted for by $\kappa_i^T$.

@@

One can also directly compare values at a specific zone due to the consistent units. For instance, zone 1 remains intermediate in capturing accessible physicians relative to zones 2 and 3 across scenarios, similar to the unconstrained case. However, the difference between travel behaviour scenarios differ in direction. Specifically, Zone 1 captures `r tot_acc$V_tot_i_2[1]- tot_acc$V_tot_i_1[1]` more and `r tot_acc$V_tot_i_1[1]- tot_acc$V_tot_i_3[1]` fewer opportunities than the lesser decay scenarios $f_2(c_{ij})$ and $f_3(c_{ij})$ respectively. Why? $\kappa_i^T$ ensures proportional allocation for each travel behaviour scenario. Meaning, while the unconstrained accessibility increases, $\kappa_i^T$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region). As the decay behaviour decreases, more opportunities are accessible for all zones. In the medium decay scenario $f_2(c_{ij})$, zone 1 sees a slight increase in values (relative to the highest decay scenario) as the zone can accessible more opportunities relative to increases seen in other zones. However, in the lowest decay scenario, zone 1 sees a decrease, as it is outpaced by increases in other zones - namely zone 2 (recall: zone 2 has the lowest number of opportunities, hence the increases in opportunity gains is much higher in a low decay scenario).

@@

Where the following $\hat \kappa_i^T$ proportional allocation factor is dimensionless:

$$
\hat \kappa_i^T = \frac{\sum_i O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
$$

@@

```{r}
#| label: tbl-simple-example-total-population-accessibility
#| tbl-cap: "Simple system: total opportunity constrained accessibility."

@@

$$
\kappa^S_{i} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
$$ {#eq-opportunity-constrained-proportional-allocation-factor}

@@

$$
V^S_{i} = \kappa^S_{i} \sum_j   D_j
$$ {#eq-attraction-constrained-accessibility-with-proportional-allocation-factor}

@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^S_{i}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$. As in the total constrained opportunity case, $\kappa^S_{i}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$. The singly constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached *and* are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered. These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of @shen1998 and @luo2003:

@@

```{r}
#| label: tbl-simple-example-attraction-constrained-accessibility
#| tbl-cap: "Simple system: singly constrained opportunity accessibility."

@@

Imposing the single proportional allocation factor $\kappa^S_{i}$ allows for the comparison of differences and ratios of the accessibility values, like previously discussed in the total constrained accessible opportunities case. The proportional allocation factor ensures that resulting values are in units of _physicians_, with the impedance units already accounted for in the allocation process.

However, unlike the total constrained opportunity case, $\kappa^S_{i}$ reflects zonal competition based on the mass of the origin, i.e., population. Again, consider the highest decay scenario $f_1(c_{ij})$. Under this scenario, zone 1 no longer captures a medium amount of physicians as in the total constrained opportunity case: it now captures the fewest in the region i.e., `r opp_acc$V_opp_i_1[1]` at zone 1, `r opp_acc$V_opp_i_1[2]` at zone 2, and `r opp_acc$V_opp_i_1[3]` at zone 3. Why may zone 1 (healthcare cluster at the edge of the city) capture `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()` of the physicians regionally while this same zone captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` in the total opportunity constrained case?

This difference is due to the single-opportunity factor $\kappa^S_{i}$'s role in $V_i^S$. The only inputs required in the total-opportunity factor $\kappa^T_i$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance. Opportunities are allocated to $i$s, regardless of the mass weights of origin. Whereas the single opportunity constraint $\kappa^S_{i}$ requires the population at $i$ as an input. In fact, $\kappa^S_{i}$ is calculated as the proportion of impedance-weighted population at an $i$ to the sum of impedance-weighted population for the entire region. Hence, since zone 1 has the lowest population in the region, is in close proximity to a more populated zone (zone 3, the 'urban core'), and is not well connected (in terms of travel impedance) to other zones with opportunities, $V_{1}^S$ values, or the number of physicians accessible, is lowest.

@@

```{r}
#| label: tbl-simple-example-attraction-constrained-accessibility-per-capita
#| tbl-cap: "Simple system: singly constrained opportunity accessibility per capita."

@@

$$
M^S_{j} = \hat \kappa^S_{j} \sum_i  O_i
$$ {#eq-production-constrained-accessibility-with-proportional-allocation-factor}

@@

$$
\kappa^S _{j} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
$$ {#eq-attraction-constrained-proportional-allocation-factor}

@@

$$
V^D_{i} = \kappa^D_{ij} \sum_j D_j
$$ {#eq-doubly-constrained-accessibility-w-paf}

@@

```{r}
#| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values
#| tbl-cap: "Doubly constrained opportunity accessibility assuming the most decay travel behaviour scenario in the modified simple system."

@@

```{r}
#| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2
#| tbl-cap: "Doubly constrained opportunity accessibility for all travel behaviour scenarios for Zone 2 in the modified simple system."

>>>>

---
title: "A family of accessibility measures derived from spatial interaction principles"
#subtitle: "Submitted preprint for peer-review (May 2025)"
author:
  - name: Anastasia Soukhov
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Rafael H. M. Pereira
    email: rafael.pereira@ipea.gov.br
    affiliations:
        - name: Institute for Applied Economic Research - Ipea
          city: Brasília, Brazil
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliations:
        - name: University of Toronto Scarborough
          department: Department of Human Geography
          city: Toronto, Canada
    attributes:
        corresponding: true
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format:
  pdf:
    keep-tex: true
    documentclass: article
    header-includes: |
      \usepackage{float}
      \floatplacement{table}{H}
      \floatplacement{figure}{H}
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
#includes:
#      in_header: 'gt_packages.sty'
      #\usepackage{float}

@@

@@

Understanding $S_i$ as the inverse of Wilson's $A_i$ does not uncover any new meaning for $S_i$ itself. Indeed, mathematically it is true, $A_i$'s role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ i.e., keeping units balanced and proportionality based on constraints. Understanding $S_i$ itself as a balancing factor is not wholly helpful as Hansen and (Stewart before him) defined accessibility as a partial sum of the demographic force $F$ or the system-wide population potential of opportunities for spatial interaction (i.e., @eq-stewart-population-potential-sum representing $V_i = \sum_j \frac{M_j}{d_{ij}}$ after losing $G$).

@@

Firstly, @tbl-small-system-land-use summarises the population (in units of 10,000s of people) and the opportunities (the number of physicians) per zone. For reference, considering @tbl-small-system-land-use's values, the Provider-to-Population-Ratio (PPR) in this system is `r sum(LU$D_j) / sum(LU$O_i)`. The number of physicians per 10,000 in Canada in 2022 was 24.97 [@whoMedicalDoctors102025]. Secondly, to pair the zonal population and opportunity information, the assumed cost of movement (in minutes of travel time) between origins and destinations is as shown in @tbl-small-system-cost.

@@

Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the proportional allocation factor for the total constrained opportunity case $\kappa_{ij}^T$ is dimensionless:

$$
\kappa_{ij}^T = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
$$

\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \sum_j \kappa_{ij}^T W^{(2)}_j$). The role of $\kappa_{ij}^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so they represent a proportion of the total number of opportunities in the region. $\kappa_{ij}^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$. This is why we refer to $\kappa_{ij}^T$ as a proportional allocation factor. On the other hand, the proportionality constant $K^T$ balances the units of $V^0_i$, the Hansen-type accessibility values, and is an alternative expression of the total constrained accessibility measure.

@@

Using the calculated proportionality constants for all zones and multiplying them by the unconstrained accessibility value $V^0_i$, the total opportunity constrained accessibility values for all zones and different travel behaviour scenarios is presented in @tbl-simple-example-total-opportunity-accessibility. $\kappa_{ij}^T$ for each zone is not shown, but can be understood to be the unitless proportion of opportunities (of the total opportunities) allocated to each zone based on travel impedance.

```{r}
#| label: tbl-simple-example-total-opportunity-accessibility
#| tbl-cap: "Simple system: total constrained accessible opportunities."

@@

In contrast to unconstrained accessibility $V^0_i$, imposing a constraint -in this case a total opportunity constraint for this variant- allows for the comparison of differences and ratios between regions and across different travel behaviour scenarios as well. Each value is effectively in units of physicians, with the impedance units already accounted for by $\kappa_{ij}^T$.

@@

One can also directly compare values at a specific zone due to the consistent units. For instance, zone 1 remains intermediate in capturing accessible physicians relative to zones 2 and 3 across scenarios, similar to the unconstrained case. However, the difference between travel behaviour scenarios differ in direction. Specifically, Zone 1 captures `r tot_acc$V_tot_i_2[1]- tot_acc$V_tot_i_1[1]` more and `r tot_acc$V_tot_i_1[1]- tot_acc$V_tot_i_3[1]` fewer opportunities than the lesser decay scenarios $f_2(c_{ij})$ and $f_3(c_{ij})$ respectively. Why? $\kappa_{ij}^T$ ensures proportional allocation for each travel behaviour scenario. Meaning, while the unconstrained accessibility increases, $\kappa_{ij}^T$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region). As the decay behaviour decreases, more opportunities are accessible for all zones. In the medium decay scenario $f_2(c_{ij})$, zone 1 sees a slight increase in values (relative to the highest decay scenario) as the zone can accessible more opportunities relative to increases seen in other zones. However, in the lowest decay scenario, zone 1 sees a decrease, as it is outpaced by increases in other zones - namely zone 2 (recall: zone 2 has the lowest number of opportunities, hence the increases in opportunity gains is much higher in a low decay scenario).

@@

Where the following $\hat \kappa_{ji}^T$ proportional allocation factor is dimensionless:

$$
\hat \kappa_{ji}^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
$$

@@

```{r}
#| label: tbl-simple-example-total-population-accessibility
#| tbl-cap: "Simple system: total constrained accessible population."

@@

$$
\kappa^S_{ij} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
$$ {#eq-opportunity-constrained-proportional-allocation-factor}

@@

$$
V^S_{i} = \sum_j \kappa^S_{ij} D_j
$$ {#eq-attraction-constrained-accessibility-with-proportional-allocation-factor}

@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^S_{ij}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$. As in the total constrained opportunity case, $\kappa^S_{ij}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$. The singly constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached *and* are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered. These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of @shen1998 and @luo2003:

@@

```{r}
#| label: tbl-simple-example-attraction-constrained-accessibility
#| tbl-cap: "Simple system: singly constrained accessible opportunities."

@@

Imposing the single proportional allocation factor $\kappa^S_{ij}$ allows for the comparison of differences and ratios of the accessibility values, like previously discussed in the total constrained accessible opportunities case. The proportional allocation factor ensures that resulting values are in units of _physicians_, with the impedance units already accounted for in the allocation process.

However, unlike the total constrained opportunity case, $\kappa^S_{ij}$ reflects zonal competition based on the mass of the origin, i.e., population. Again, consider the highest decay scenario $f_1(c_{ij})$. Under this scenario, zone 1 no longer captures a medium amount of physicians as in the total constrained opportunity case: it now captures the fewest in the region i.e., `r opp_acc$V_opp_i_1[1]` at zone 1, `r opp_acc$V_opp_i_1[2]` at zone 2, and `r opp_acc$V_opp_i_1[3]` at zone 3. Why may zone 1 (healthcare cluster at the edge of the city) capture `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()` of the physicians regionally while this same zone captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` in the total opportunity constrained case?

This difference is due to the single-opportunity factor $\kappa^S_{ij}$'s role in $V_i^S$. The only inputs required in the total-opportunity factor $\kappa^T_{ij}$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance. Opportunities are allocated to $i$s, regardless of the mass weights of origin. Whereas the single opportunity constraint $\kappa^S_{ij}$ requires the population at $i$ as an input. In fact, $\kappa^S_{ij}$ is calculated as the proportion of impedance-weighted population at an $i$ to the sum of impedance-weighted population for the entire region. Hence, since zone 1 has the lowest population in the region, is in close proximity to a more populated zone (zone 3, the 'urban core'), and is not well connected (in terms of travel impedance) to other zones with opportunities, $V_{1}^S$ values, or the number of physicians accessible, is lowest.

@@

```{r}
#| label: tbl-simple-example-attraction-constrained-accessibility-per-capita
#| tbl-cap: "Simple system: singly constrained accessible opportunities per capita."

@@

$$
M^S_{j} = \hat \kappa^S_{ji} \sum_i  O_i
$$ {#eq-production-constrained-accessibility-with-proportional-allocation-factor}

@@

$$
\hat \kappa^S_{ji} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
$$ {#eq-attraction-constrained-proportional-allocation-factor}

@@

$$
V^D_{i} = \sum_j \kappa^D_{ij} D_j
$$ {#eq-doubly-constrained-accessibility-w-paf}

@@

```{r}
#| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values
#| tbl-cap: "Doubly constrained accessible opportunities assuming highest travel decay in the modified simple system."

@@

```{r}
#| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2
#| tbl-cap: "Doubly constrained accessible opportunities at Zone 2 for all travel decay groups in the modified simple system."
